Recurrent event analysis in the presence of real-time high frequency data via random subsampling

### 2019-07-08 write-up/fda-recurrent.tex
Adding multiple imputation section

A core concern in mHealth is that sensor data can be missing due to
sensor wearing issue.
To see this, for the suicidal ideation case study, there are 2139
self-identified moments of distress across all 91 participants.  Of
these, 1289 event times had complete data for the prior thirty minutes,
1984 had fraction of missing data on a fine grid less than 30\%, and
1998 had fraction of missing data on a fine grid less than 10\%.

Missing data is a critical issue because~$c_{i,k} (t)$ cannot be
estimated if~$X(s,t)$ is not observed for all~$s \in [0,\Delta]$. 
Moreover, we want standard errors to reflect the uncertainty
in these coefficients when missing data is prevalent.
\cite{Goldsmith2015} suggest using best linear unbiased predictors
(BLUP) or posterior modes in the mixed effects model to
estimate~$c_{i,k} (t)$; however, one criticism would be when there is
substantial variability in these estimates.  To deal with this,
\cite{Crainiceanu2010, Goldsmith2010} take a full Bayesian analysis.
\cite{Yao2005} introduced PACE as a frequentist analysis that deals
with this issue.
However,~\cite{Reimherr2018} show that for sparse, irregular
longitudinal, the imputation model should not ignore the outcome~$Y_i
(t)$.

Here we present an extension of~\cite{Reimherr2018} to our setting,
leverage Lemma~\ref{lemma:logistic} and the marginal covariance
estimation procedure to build a multiple imputation approach.
\[
\E [ X_i (s,t) \given Y_i(t) = y, \bfx ] = \mu_y (s,t) + {\bf a}_{i,t}^\top
{\bf B}_i (\bfx_i (t) - \mathbold{\mu}_i )
\]
where~$\mu_y (t) = \E [ X_i (t) \given Y_i (t) = y]$, and
\[
{\bf a}_i^\top = ( \Sigma(s, )
\]
This generalizes the procedure of~\cite{Reimerr2018} to the
longitudinal data setting.
This is possible due to  and our 
Here, we ignore the issue that~$X(t,s)$ is

The estimate is then given by~$\bar \theta = M^{-1} \sum_{m=1}^M \hat
\theta^{(m)}$; the covariance matrix can be estimated by
\[
M^{-1} \sum_{m=1}^M \hat \Xi^{(m)} (\theta) + \frac{1}{M-1}
\sum_{m=1}^M \hat \theta - \bar \theta) (\hat \theta - \bar \theta)
\]
where~$\hat \Xi^{(m)}$ is the estimator~\eqref{eq:fisher} for
the~$m$th imputated dataset.

@@

\subsection{Re-using estimates under additional sampling}

If~$\pi$ is constant then we can continue to use the same pooled
covariance.

\section{Simulations}

\subsection{Unbiased estimation}

@@

@@

\section{Case Study: }

>>>>

A core concern in mHealth is that sensor data can often be missing for
intervals of time due to sensor wearing issue (e.g., sensor band
becomes loose, turns off accidentally, data packets lost, or battery
dies). In the suicidal ideation case study, for example, there are
2139 self-identified moments of distress across all 91 participants.
Of these, 1289 event times had complete data for the prior thirty
minutes, 1984 had fraction of missing data on a fine grid less than
30\%, and 1998 had fraction of missing data on a fine grid less than
10\%.

Missing data is a critical issue because~$c_{i,k} (t)$ cannot be
estimated if~$X(s,t)$ is not observed for all~$s \in [0,\Delta]$. 
Moreover, we want standard errors to reflect the uncertainty
in these coefficients when missing data is prevalent.
\cite{Goldsmith2011} suggest using best linear unbiased predictors
(BLUP) or posterior modes in the mixed effects model to
estimate~$c_{i,k} (t)$; however, one criticism would be when there is
substantial variability in these estimates.  To deal with this,
\cite{Crainiceanu2010, Goldsmith2010} take a full Bayesian analysis.
\cite{Yao2005} introduced PACE as an alternative frequentist method.
However,~\cite{Petrovich2018} shows that for sparse, irregular
longitudinal, the imputation model should not ignore the outcome
variable~$Y_i (t)$.

Here we present an extension of~\cite{Petrovich2018} to our setting by
leveraging Lemma~\ref{lemma:logistic} and the marginal covariance
estimation procedure to construct a multiple imputation procedure.
Let~$\bfx_{i} (t)$ denote incomplete sensor data at time~$t$ (i.e., at
times~$\{ s_{i,r} \}_{r=1}^{k_{it}} \}$ in~$[0,\Delta]$. Then
\begin{align}
\label{eq:impute}
\E [ X_i (s,t) \given Y_i(t) = y, \bfx_i (t) ] 
  &= \mu_y (s,t) + {\bf a}_{i,t}^\top (s) {\bf B}_{i,t} (\bfx_i (t) -
    \mathbold{\mu}_i (t) ) \\
\text{Cov} \left( X_i (s,t), X_i (s^\prime, t) \given Y_i (t) = y,
  \bfx_i (t) \right) 
  &= \Sigma_{t} (s, s^\prime) - 
    {\bf a}_{i,t} (s)^\top {\bf B}_{i,t} {\bf a}_{i,t} (s^\prime)
\end{align}
where we have
\[
{\bf a}_{i,t} (s)^\top = \left( \begin{array}{c} \Sigma_t (s_{i,1}, s) \\ \vdots \\
                       \Sigma_t (s_{i,k_{it}}, s) \end{array} \right); \quad
{\bf B}_{i,t}^{-1} = \left( 
  \begin{array}{c c c} 
    \Sigma_t(s_{i,1}, s_{i,1}) & \Sigma_t(s_{i,1}, s_{i,2}) & \cdots \\
    \Sigma_t(s_{i,1}, s_{i,2}) & \ddots & \vdots \\
    \vdots & \cdots & \Sigma_t (s_{i,k_{it}}, s_{i,k_{it}}) 
  \end{array} \right),
\]
$\mathbold{\mu}_i (t) = \E [ \bfx_i (t) \given Y_i (t) = y] = \{
\mu_y (s_{j}, t) \}_{j=1}^r$ and~$\mu_y (s,t)$ is the mean of~$X(t,s)$
from group~$y$, and~$\Sigma_t ( s^\prime, s)$ is the covariance
between~$X(s^\prime,t)$ and $X(s,t)$ for $s^\prime, s \in [0, \Delta]$ 
and~$t \in \mathbb{R}_+$.
These covariances can be estimated using the methods considered by
\cite{ChenMuller2012}.  Our imputation procedure instead uses the
marginal covariance~$\Xi$ as a proxy for the true covariance
function~$\Sigma_t(\cdot, \cdot)$.  We found this to work well in
simulations as~$\Sigma_t$ was relatively constant across~$t$; however,
this imputation strategy may be inappropriate when the covariance
structure changes significantly with time~$t$.

The imputation strategy generalizes the procedure introduced
by~\cite{Petrovich2018} to the longitudinal functional data
setting.  This is only possible due to the connection established with
logistic regression by Lemma~\ref{lemma:logistic}.
We consider a multiple imputation estimate with~$M$ imputations;
the likelihood estimate is given by~$\bar \theta = M^{-1}
\sum_{m=1}^M \hat \theta^{(m)}$ where~$\theta^{(m)}$ is the estimate
from the~$m$th imputated dataset; the covariance matrix can be
estimated by 
\[
M^{-1} \sum_{m=1}^M \hat \Xi^{(m)} (\theta) + \frac{1}{M-1}
\sum_{m=1}^M (\hat \theta^{(m)} - \bar \theta) (\hat \theta^{(m)} -
\bar \theta)
\]
where~$\hat \Xi^{(m)}$ is the estimator~\eqref{eq:fisher} for
the~$m$th imputated dataset.

@@

\section{Simulation study}

We next assess the proposed methodology via a simulation study.
Simulation studies of joint high frequency sensor and recurrent
event data can be quite difficult to construct.  Here, we use the
imputation strategy introduced in Section~\ref{section:missingdata} to
circumvent some of the issues.  
In this section, we assume each individual is observed over five
12-hour days.

First, we generate a sequence of event times via a marginal hazard
rate that ignores sensor data.

We generate 5 days of 12 hours with range of marginal rates (1 per
12 hours, 1 per 6 hours, 1 per hour).  We then sample non-events at
rates or (1 per 12/6/1/0.5 hours).
We assume that~$Y_t = 0$ implies the curve is flat with mean 0.
We assume that~$Y_t = 1$ implies three curves (A) the curve jumps
around 10 minutes prior to the event and remains elevated, (B) the
curve goes down then up and down smoothly, (C) 
These curves are given in Figure~\ref{} for three choices of
parameters that indicate varying degrees of departure from the~$Y_t =
0$ case.

Using~\eqref{eq:}
We define~$X(t,s)$ at the grid of observerations per secondw ith
\[
  \Sigma_t (s, s^\prime) = 
  \frac{\sigma^2}{\Gamma (v) 2^{\nu-1}} 
  \left( \frac{\sqrt{2 \nu} |t-s|}{\rho_t} \right)^{\nu}
  K_v \left( \frac{\sqrt{2 \nu} |t-s|}{\rho_t} \right)
\]
where~$K_v$ is the modified Bessel function of the second kind, and
we choose~$\nu = 5/2$ and $\sigma^2 = 1$. 
We choose three versions of~$\rho_t$: (A) constant across time
at $0.5$, (B) constant within each day but linearly decreasing
across days from~$0.7$ to~$0.3$, and (C) constant within each day but
linearly increasing across days from~$0.3$ to~$0.7$.

These curves are evaluate

We estimate with~$K_b = K_x = 35$ under both the logistic, penalized
logistic, and multi-level penalized logistic models.

\subsection{Mean-integrated squared error (MISE)}

Since we are primarily interested in accuracy of the final
estimates~$\hat \beta(s)$, we report the mean integrated  squared
error (MISE) of~$\hat \beta (s)$, defined as~$\frac{1}{1000}
\sum_{j=1}^{1000} \int_{-\infty}^t ( \hat \beta_j (t) - \beta(s))^2
ds$ where~$\beta(s) \equiv 0$ for $s > \Delta$.  We define the MISE in
this manner to account for settings where~$\Delta$ is unknown.
For~$\Delta$ known, the MISE simplifies to integrals from~$t-\Delta$
to~$t$.
Table~\ref{} show the MISE of~$\hat \beta (s)$ decomposed into the
variance and squared biase

@@

In each (A)-(C) we parametrize the curves and check the type I error
for various sizes of the

@@

\section{A worked example: Adolescent psychiatric inpatient mHealth study}

During a XX month period in 2019, 100 psychiatric inpatients admitted
for suicidal risk to Franciscan Children's Hospital

hospitals with histologically verified liver cirrhosis were recu
those presenting to a psychiatric hospital with suicide ideation
and/or a recent suicide attempt) during a high risk time period (i.e.,
post-hospitalization)

Currently, only 51 patients for whom the initial biopsy could be
reevaluated using more restrictive criteria were available for
analysis.
The goal of the study was to ask whether sensor information can used
to identify distress early prior in the process of escalation by
monitoring physiological correlates of distress -- in particular,
electrodermal activity (EDA), a measure of skin conductance, and
``gACC'', a coordinate-free feature built from accelerometer data 
that measures physical movement. 
The scientific hypothesis to be tested is whether a spike in
EDA and/or gACC

The goal of the sutyd

EDA and ACC sensor data along with buttonpress timestamps were
graciously provided by Evan Kleiman and his study team {\tt
https://kleimanlab.org}.
At the end of the study period, the mortality rate was 292/488, or
approximately 60\%.

The focus here is on two EDA and the  prothrombin index, a composite
blood coagulation index
related to liver function, measured initially at three-month intervals
and subsequently at roughly twelve-month intervals.
The individual prothrombin trajectories are highly variable,
both in forward and in reverse time,
which tends to obscure patterns and trends.
In Figure~\ref{fig:mean_traj}a the mean trajectory is plotted against time from recruitment
for two patient groups placebo/prednisone and censored/not censored.
Naturally, only those patients who are still alive are included in the average
for that time.
Figure~\ref{fig:mean_traj}b shows the same plots in reverse alignment.
While there are certain similarities in the two plots,
the differences in temporal trends are rather striking.
In particular, prothrombin levels in the six months prior to censoring are fairly stable,
which is in marked contrast with levels in the six months prior to failure,
as seen in the lower pair of curves.

Inspection of the graphs for uncensored patients in the right panel of Figure~\ref{fig:mean_traj} suggests
beginning with the simplest revival model in which the sequences
for distinct patients are independent Gaussian with moments

\subsection{Effect of electrodermal activity on prognosis}
Over a period of 5~years and one month following recruitment,
patient~$u$ had eight appointments with prothrombin values as follows:
